### 2016-09-21 untitled.Rmd
Untitled

---
title: "Untitled"
authors: Ben Marwick, Maria Schaarschmidt 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2016-09-27 analysis/paper/untitled.Rmd
add outline

---
title: 'The title'
author:
- Ben Marwick
- Maria Schaarschmidt
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   bookdown::word_document2:
    fig_caption: yes
    # reference_docx: templates/template.docx
bibliography: references.bib
# csl: journal-of-archaeological-science.csl
abstract: |
  This is the abstract.
---
```{r setup, include=FALSE, echo = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE)
opts_chunk$set(echo = FALSE, 
               messages = FALSE, 
               warnings = FALSE, 
               dpi = 300,
               cache = TRUE)
library(tidyverse)
```

# Introduction

# Background

## Geomorphology of the Irrawady terraces

- survey early literature on red earth deposits

## Archaeology of the Irrawady terraces

- Movius and contemporaries
- Dennell is a major commentator recently

## Survey data

```{r}

```

- our landscape observations
- descriptions of existing cuttings that we visited
- GPS data from transect
- surface finds of artefacts by KK

## Formation of the terraces and archaeological contexts

- summary of formation 
- implications for age of the archaeological deposit

## Conclusion

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

In Table \@ref(tab:carstab) we can see some data about the relationship between pressure and tempurature.

```{r carstab}
library(knitr)
kable(head(mtcars), caption = "Data about cars")
```

## Including Plots

You can also embed plots, for example:

```{r pressureplot, echo=FALSE, fig.cap="Plot of car data"}
plot(pressure)
```

In Figure \@ref(fig:pressureplot) we can see some data about pressure.

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Citations and References

And we can have a citation, using a `.bib` file that holds all the details. To get this: [@Marwick2016repro] we type `[@Marwick2016repro]`. The text after the `@` is the bibtex key that links the in-text citation to the full details in the `.bib` file.

All of the usual variations on in-text citation formatting are possible in markdown, and listed for reference here: <http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html>

## Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
# which R packages and versions?
kable(data.frame(Setting = names(devtools::session_info()$platform),
                 Value = sapply(devtools::session_info()$platform, `[`, 1),
                 stringsAsFactors = FALSE,
                 row.names = NULL),
      caption = "R session information")
```

```{r}
kable(devtools::session_info()$packages,
      caption = "Packages that this report depends on")

# what commit is this file at?
library(git2r)
repo <- repository(path = "../..")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is "`r last_commit@summary`".

## References

### 2016-10-03 analysis/paper/paper.Rmd
update pkgs
regional structural geology

--- 2016-09-27 analysis/paper/untitled.Rmd
@@
 
+The study area is located between the Sagaing fault to the west and the Shan scarp to the east.
+The Sagaing fault is a major strike-slip right-lateral continental fault that extends over 1200 km, first described by Noetling [-@noetling1900miocene], and later described by others [e.g.
+@thein1991lateral; maung1987transcurrent].
+The Shan scarp is a topographic discontinuity that marks the boundary of the central plains and the Shan plateau to the east, a region with an average elevation of 1000 m and large variations of elevation over short distances (up to 1800 m over a few kilometers) [@bertrand2001diachronous; bertrand2003tectonics].
+
 ## Geomorphology of the Irrawady terraces
@@
 ```{r}
-
+# analysis of survey and GPS data in here
 ```

### 2016-10-03 analysis/paper/paper.Rmd
First draft (comments of Dennell and Problems with Movius Line)

@@
 - Movius and contemporaries
-- Dennell is a major commentator recently
+
+## Dennell is a major commentator recently
+Robin Dennell is one of the more recent critics of the concept of the Movius Line.
+In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” (Dennell, 2016).
+The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution.
+According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.
+Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia (Dennell, 2016).
+In contrast to Robin Dennell’s views, Norton et al.
+(2006) reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line.
+The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts (Norton et al., 2006; Norton & Bae, 2008).
+Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary (Dennell, 2016).
+Reasons for the lack of Acheulean technology are investigated in modern quaternary science (Brumm & Moore, 2012; Dennell, 2016; Norton & Bae, 2008; Norton, Bae, Harris, & Lee, 2006; Petraglia & Shipton, 2008; Schick, 1994), including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
+The use of stone tools may have been influenced by certain situations, tasks or individuals.
+Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times.
+Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool (Dennell, 2016).
+
+Raw Material
+
+Movius (1944) stated that the raw material used east of the Movius Line was often of low quality quartz and quartzite that would prevent the production of the same kind of bifaces (Dennell, 2016; Lycett & Bae, 2010).
+However, handaxes from Zhoukoudian Locality 1, Chongokni and Kumpari (China) are bifacially worked and are made of quartz and quartzite river cobbles from this area (Lycett & Bae, 2010).
+
+Environmental Changes
+
+Whilst colonizing East Asia, early modern humans encountered barriers such as mountain ranges, river deltas, oceans and deserts that provide a range of resources (e.g.
+water, food, shelter) (Lycett & Bae, 2010).
+However, the environment during the Middle Pleistocene was challenging in terms of changing climatic conditions and accompanying biogeographic transitions (Bar-Yosef & Belfer-Cohen, 2001; Field & Lahr, 2006; Lycett & Bae, 2010; Schick, 1994).
+
+There may have been no need for intensive stone tool production, or a different material was used at some point.
+This theory is known as the “bamboo hypothesis” (Field & Lahr, 2006; Lycett & Bae, 2010; Schick, 1994) and states that early modern humans used bamboo instead of stones to make tools (Schick, 1994) leaving behind no archaeological record (Lycett & Bae, 2010).
+However, even though bamboo cutting marks on bones can be distinguished from other cutting marks, there have been no findings of such so far (Lycett & Bae, 2010).
+
+Dispersal routes
+
+The dispersal route of early modern humans from Africa to Asia and Australia is still debated among quaternary scientists.
+Based on GIS analyses, Field & Lahr (2006) and Field et al.
+(2007) identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia.
+This “Southern Dispersal Route” is dated to around 75-60ka (Field & Lahr, 2006; Field, Petraglia, & Lahr, 2006; Clarkson et al., 2011; Macaulay et al., 2005).
+This route matches with some Middle Pleistocene archaeological sites in Africa, India (e.g.
+Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions.
+Of particular importance are the deposits at the Hiran Valley since they have been dated to 56-69ka, matching the proposed time for the Southern Dispersal.
+A dispersal along the coasts and rivers seems reasonable because they provide sufficient resources.
+However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes QUELLE.
+
+Demographic and social transmission
+
+Recently, the model of demographic and social transmission has been proposed as a possible explanation for the diverse development of technologies in E/SE Asia and W/SW Asia.
+The concept is based on the assumption that a certain effective population size is essential for developing traditions and technologies to be passed on to further generations (Lycett & Bae, 2010; Lycett & Norton, 2010; Lycett, 2007).
+Given the colonization of the whole of East Asia and the distance to Africa decreasing the population size, the population density may have been too low to maintain or establish more elaborate tool making techniques (Lycett & Norton, 2010; Lycett, 2007).
+[Figure demographic and social transmission?]
+
+Definition of “Acheulean”
+
+Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma (Brumm & Moore, 2012; Lycett & Bae, 2010; Norton et al., 2006).
+In contrary to Mode I technology, these stone tools show a degree of standardization (Norton et al., 2006) and were the main tools used by hominins during the Pleistocene (Petraglia & Shipton, 2008).
+The tools include handaxes, cleavers, picks, knives, lanceolates and unifaces (Brumm & Moore, 2012).
+Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well (Brumm & Moore, 2012).
+The definition of the Acheulean is not always straightforward, i.e.
+the assignment of finds can be influenced by the finding location east or west of the Movius Line QUELLE.
+Also, bifaces are not suitable as a time marker for specific cultures as they have been produced in different parts of the world over several 100 ka (Dennell, 2016).
 
@@
 
-The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`.
-The current commit message is "`r last_commit@summary`".
-
 ## References

### 2016-10-03 analysis/paper/paper.Rmd
deleted Robin

@@
 
-## Dennell is a major commentator recently
-Robin Dennell is one of the more recent critics of the concept of the Movius Line.
-In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” (Dennell, 2016).
+# Dennell is a major commentator recently
+Dennell is one of the more recent critics of the concept of the Movius Line.
+In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
 The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution.
@@
 Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia (Dennell, 2016).
-In contrast to Robin Dennell’s views, Norton et al.
+In contrast to Dennell’s views, Norton et al.
 (2006) reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line.
 The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts (Norton et al., 2006; Norton & Bae, 2008).
-Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary (Dennell, 2016).
-Reasons for the lack of Acheulean technology are investigated in modern quaternary science (Brumm & Moore, 2012; Dennell, 2016; Norton & Bae, 2008; Norton, Bae, Harris, & Lee, 2006; Petraglia & Shipton, 2008; Schick, 1994), including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
+Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life].
+Reasons for the lack of Acheulean technology are investigated in modern quaternary science (Brumm & Moore, 2012; @dennell2016life; Norton & Bae, 2008; Norton, Bae, Harris, & Lee, 2006; Petraglia & Shipton, 2008; Schick, 1994), including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
 The use of stone tools may have been influenced by certain situations, tasks or individuals.
 Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times.
-Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool (Dennell, 2016).
+Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool [@dennell2016life].
 

### 2016-10-04 analysis/paper/paper.Rmd
added references

@@
 The Sagaing fault is a major strike-slip right-lateral continental fault that extends over 1200 km, first described by Noetling [-@noetling1900miocene], and later described by others [e.g.
-@thein1991lateral; maung1987transcurrent].
-The Shan scarp is a topographic discontinuity that marks the boundary of the central plains and the Shan plateau to the east, a region with an average elevation of 1000 m and large variations of elevation over short distances (up to 1800 m over a few kilometers) [@bertrand2001diachronous; bertrand2003tectonics].
+@thein1991lateral; @maung1987transcurrent].
+The Shan scarp is a topographic discontinuity that marks the boundary of the central plains and the Shan plateau to the east, a region with an average elevation of 1000 m and large variations of elevation over short distances (up to 1800 m over a few kilometers) [@bertrand2001diachronous; @bertrand2003tectonics].
 
@@
 According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.
-Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia (Dennell, 2016).
+Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].
 In contrast to Dennell’s views, Norton et al.
 (2006) reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line.
-The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts (Norton et al., 2006; Norton & Bae, 2008).
+The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts [@norton2006middle; @norton2008movius].
 Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life].
-Reasons for the lack of Acheulean technology are investigated in modern quaternary science (Brumm & Moore, 2012; @dennell2016life; Norton & Bae, 2008; Norton, Bae, Harris, & Lee, 2006; Petraglia & Shipton, 2008; Schick, 1994), including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
+Reasons for the lack of Acheulean technology are investigated in modern quaternary science [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius], including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
 The use of stone tools may have been influenced by certain situations, tasks or individuals.
@@
 
-Movius (1944) stated that the raw material used east of the Movius Line was often of low quality quartz and quartzite that would prevent the production of the same kind of bifaces (Dennell, 2016; Lycett & Bae, 2010).
-However, handaxes from Zhoukoudian Locality 1, Chongokni and Kumpari (China) are bifacially worked and are made of quartz and quartzite river cobbles from this area (Lycett & Bae, 2010).
+Movius (1944) stated that the raw material used east of the Movius Line was often of low quality quartz and quartzite that would prevent the production of the same kind of bifaces [@dennell2016life; @lycett2010movius].
+However, handaxes from Zhoukoudian Locality 1, Chongokni and Kumpari (China) are bifacially worked and are made of quartz and quartzite river cobbles from this area [@lycett2010movius].
 
@@
 Whilst colonizing East Asia, early modern humans encountered barriers such as mountain ranges, river deltas, oceans and deserts that provide a range of resources (e.g.
-water, food, shelter) (Lycett & Bae, 2010).
-However, the environment during the Middle Pleistocene was challenging in terms of changing climatic conditions and accompanying biogeographic transitions (Bar-Yosef & Belfer-Cohen, 2001; Field & Lahr, 2006; Lycett & Bae, 2010; Schick, 1994).
+water, food, shelter) [@lycett2010movius].
+However, the environment during the Middle Pleistocene was challenging in terms of changing climatic conditions and accompanying biogeographic transitions [@bar2001africa; @field2005assessment; @lycett2010movius; @schick1994movius].
 
 There may have been no need for intensive stone tool production, or a different material was used at some point.
-This theory is known as the “bamboo hypothesis” (Field & Lahr, 2006; Lycett & Bae, 2010; Schick, 1994) and states that early modern humans used bamboo instead of stones to make tools (Schick, 1994) leaving behind no archaeological record (Lycett & Bae, 2010).
-However, even though bamboo cutting marks on bones can be distinguished from other cutting marks, there have been no findings of such so far (Lycett & Bae, 2010).
+This theory is known as the “bamboo hypothesis” [@field2005assessment; @lycett2010movius; @schick1994movius] and states that early modern humans used bamboo instead of stones to make tools [@schick1994movius] leaving behind no archaeological record [@lycett2010movius].
+However, even though bamboo cutting marks on bones can be distinguished from other cutting marks, there have been no findings of such so far [@lycett2010movius].
 
@@
 (2007) identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia.
-This “Southern Dispersal Route” is dated to around 75-60ka (Field & Lahr, 2006; Field, Petraglia, & Lahr, 2006; Clarkson et al., 2011; Macaulay et al., 2005).
+This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single].
 This route matches with some Middle Pleistocene archaeological sites in Africa, India (e.g.
@@
 Recently, the model of demographic and social transmission has been proposed as a possible explanation for the diverse development of technologies in E/SE Asia and W/SW Asia.
-The concept is based on the assumption that a certain effective population size is essential for developing traditions and technologies to be passed on to further generations (Lycett & Bae, 2010; Lycett & Norton, 2010; Lycett, 2007).
-Given the colonization of the whole of East Asia and the distance to Africa decreasing the population size, the population density may have been too low to maintain or establish more elaborate tool making techniques (Lycett & Norton, 2010; Lycett, 2007).
+The concept is based on the assumption that a certain effective population size is essential for developing traditions and technologies to be passed on to further generations [@lycett2010movius; @lycett2010demographic; @lycett2007there].
+Given the colonization of the whole of East Asia and the distance to Africa decreasing the population size, the population density may have been too low to maintain or establish more elaborate tool making techniques [@lycett2010demographic; @lycett2007there].
+
 [Figure demographic and social transmission?]
@@
 
-Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma (Brumm & Moore, 2012; Lycett & Bae, 2010; Norton et al., 2006).
-In contrary to Mode I technology, these stone tools show a degree of standardization (Norton et al., 2006) and were the main tools used by hominins during the Pleistocene (Petraglia & Shipton, 2008).
-The tools include handaxes, cleavers, picks, knives, lanceolates and unifaces (Brumm & Moore, 2012).
-Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well (Brumm & Moore, 2012).
+Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma [@brumm2012biface; @lycett2010movius; @norton2006middle].
+In contrary to Mode I technology, these stone tools show a degree of standardization [@norton2006middle] and were the main tools used by hominins during the Pleistocene [@petraglia2008large].
+The tools include handaxes, cleavers, picks, knives, lanceolates and unifaces [@brumm2012biface].
+Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well [@brumm2012biface].
 The definition of the Acheulean is not always straightforward, i.e.
 the assignment of finds can be influenced by the finding location east or west of the Movius Line QUELLE.
-Also, bifaces are not suitable as a time marker for specific cultures as they have been produced in different parts of the world over several 100 ka (Dennell, 2016).
+Also, bifaces are not suitable as a time marker for specific cultures as they have been produced in different parts of the world over several 100 ka [@dennell2016life].
 

### 2016-10-04 analysis/paper/paper.Rmd
updated references

@@
 In contrast to Dennell’s views, Norton et al.
-(2006) reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line.
+[-@norton2006middle] reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line.
 The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts [@norton2006middle; @norton2008movius].
@@
 
-Movius (1944) stated that the raw material used east of the Movius Line was often of low quality quartz and quartzite that would prevent the production of the same kind of bifaces [@dennell2016life; @lycett2010movius].
+Movius [-@movius1944early] stated that the raw material used east of the Movius Line was often of low quality quartz and quartzite that would prevent the production of the same kind of bifaces [@dennell2016life; @lycett2010movius].
 However, handaxes from Zhoukoudian Locality 1, Chongokni and Kumpari (China) are bifacially worked and are made of quartz and quartzite river cobbles from this area [@lycett2010movius].
@@
 The dispersal route of early modern humans from Africa to Asia and Australia is still debated among quaternary scientists.
-Based on GIS analyses, Field & Lahr (2006) and Field et al.
-(2007) identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia.
+Based on GIS analyses, Field & Lahr [-@field2005assessment] and Field et al.
+[-@field2007southern] identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia.
 This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single].

### 2016-10-04 analysis/paper/paper.Rmd
intro and early geo literature

--- 2016-10-03 analysis/paper/paper.Rmd
@@
 
+The Irrawaddy River terraces of central Myanmar have been part of global Pleistocene archaeology since de Terra and Movius [-@deTerra1943] published their monograph reporting archaeological evidence from the Middle Pleistocene.
+Their expedition also resulted in Movius' [-@movius1948lower] influential paper claiming to identify a major cultural and technological division between the east and west hemispheres.
+Movius described a region to the west of the Irrawaddy where hand-axes are present in Middle Pleistocene contexts, and a region to the east where hand-axes are absent, with chopper-chopping tools found instead.
+This division has subsequently been known as the Movius Line, and has been the subject of extensive discussion by archaeologists [@lycett2010demographic; @petraglia2008large; @lycett2010movius; keates2002movius; @norton2006movius; @yamei2000mid; @norton2008movius; @brantingham1999astride; @brumm2012biface; @brumm2010movius; @norton2006middle; @davis1999recent; @norton2000current; @lycett2007soanian; Romanowska2016].
+This includes critical discussion that has questioned the reliability of Movius' observations
+
+The importance of the Movius' findings depend on the chronology of the artefacts he discovered.
+inferred from his geomorphological interpretations.
+Despite the prominence of the Movius Line in discussions of global patterns of human evolution, there has not been any new data from the location of Line since Movius' initial field work.
+In this paper we assess the current status of the
+
 # Background
@@
 
+Descriptions of the physical and cultural geography of the Irrawaddy terraces of central Myanmar have been appearing scholarly literature since Myanmar was attached to the British Empire in 1886.
+In the following section we review this literature to characterise this landscape.
+Then we review previous work on the archaeology of this landscape.
+Finally, we report on our recent surveys in the Chauk region, and discuss the implications for the archaeological significance of the region.
+
 ## Geomorphology of the Irrawady terraces
+
+Rep
+
+chhibber1934geology
 
@@
 - Movius and contemporaries
-
-# Dennell is a major commentator recently
-Dennell is one of the more recent critics of the concept of the Movius Line.
-In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
-The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution.
-According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.
-Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia (Dennell, 2016).
-In contrast to Dennell’s views, Norton et al.
-(2006) reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line.
-The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts (Norton et al., 2006; Norton & Bae, 2008).
-Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life].
-Reasons for the lack of Acheulean technology are investigated in modern quaternary science (Brumm & Moore, 2012; @dennell2016life; Norton & Bae, 2008; Norton, Bae, Harris, & Lee, 2006; Petraglia & Shipton, 2008; Schick, 1994), including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
-The use of stone tools may have been influenced by certain situations, tasks or individuals.
-Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times.
-Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool [@dennell2016life].
-
-Raw Material
-
-Movius (1944) stated that the raw material used east of the Movius Line was often of low quality quartz and quartzite that would prevent the production of the same kind of bifaces (Dennell, 2016; Lycett & Bae, 2010).
-However, handaxes from Zhoukoudian Locality 1, Chongokni and Kumpari (China) are bifacially worked and are made of quartz and quartzite river cobbles from this area (Lycett & Bae, 2010).
-
-Environmental Changes
-
-Whilst colonizing East Asia, early modern humans encountered barriers such as mountain ranges, river deltas, oceans and deserts that provide a range of resources (e.g.
-water, food, shelter) (Lycett & Bae, 2010).
-However, the environment during the Middle Pleistocene was challenging in terms of changing climatic conditions and accompanying biogeographic transitions (Bar-Yosef & Belfer-Cohen, 2001; Field & Lahr, 2006; Lycett & Bae, 2010; Schick, 1994).
-
-There may have been no need for intensive stone tool production, or a different material was used at some point.
-This theory is known as the “bamboo hypothesis” (Field & Lahr, 2006; Lycett & Bae, 2010; Schick, 1994) and states that early modern humans used bamboo instead of stones to make tools (Schick, 1994) leaving behind no archaeological record (Lycett & Bae, 2010).
-However, even though bamboo cutting marks on bones can be distinguished from other cutting marks, there have been no findings of such so far (Lycett & Bae, 2010).
-
-Dispersal routes
-
-The dispersal route of early modern humans from Africa to Asia and Australia is still debated among quaternary scientists.
-Based on GIS analyses, Field & Lahr (2006) and Field et al.
-(2007) identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia.
-This “Southern Dispersal Route” is dated to around 75-60ka (Field & Lahr, 2006; Field, Petraglia, & Lahr, 2006; Clarkson et al., 2011; Macaulay et al., 2005).
-This route matches with some Middle Pleistocene archaeological sites in Africa, India (e.g.
-Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions.
-Of particular importance are the deposits at the Hiran Valley since they have been dated to 56-69ka, matching the proposed time for the Southern Dispersal.
-A dispersal along the coasts and rivers seems reasonable because they provide sufficient resources.
-However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes QUELLE.
-
-Demographic and social transmission
-
-Recently, the model of demographic and social transmission has been proposed as a possible explanation for the diverse development of technologies in E/SE Asia and W/SW Asia.
-The concept is based on the assumption that a certain effective population size is essential for developing traditions and technologies to be passed on to further generations (Lycett & Bae, 2010; Lycett & Norton, 2010; Lycett, 2007).
-Given the colonization of the whole of East Asia and the distance to Africa decreasing the population size, the population density may have been too low to maintain or establish more elaborate tool making techniques (Lycett & Norton, 2010; Lycett, 2007).
-[Figure demographic and social transmission?]
-
-Definition of “Acheulean”
-
-Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma (Brumm & Moore, 2012; Lycett & Bae, 2010; Norton et al., 2006).
-In contrary to Mode I technology, these stone tools show a degree of standardization (Norton et al., 2006) and were the main tools used by hominins during the Pleistocene (Petraglia & Shipton, 2008).
-The tools include handaxes, cleavers, picks, knives, lanceolates and unifaces (Brumm & Moore, 2012).
-Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well (Brumm & Moore, 2012).
-The definition of the Acheulean is not always straightforward, i.e.
-the assignment of finds can be influenced by the finding location east or west of the Movius Line QUELLE.
-Also, bifaces are not suitable as a time marker for specific cultures as they have been produced in different parts of the world over several 100 ka (Dennell, 2016).
+- Dennell is a major commentator recently
 
@@
 
+The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`.
+The current commit message is "`r last_commit@summary`".
+
 ## References

### 2016-10-04 analysis/paper/paper.Rmd
more on intro

@@
 This division has subsequently been known as the Movius Line, and has been the subject of extensive discussion by archaeologists [@lycett2010demographic; @petraglia2008large; @lycett2010movius; keates2002movius; @norton2006movius; @yamei2000mid; @norton2008movius; @brantingham1999astride; @brumm2012biface; @brumm2010movius; @norton2006middle; @davis1999recent; @norton2000current; @lycett2007soanian; Romanowska2016].
-This includes critical discussion that has questioned the reliability of Movius' observations
+This includes critical discussion that has questioned the reliability of Movius' field observations [@dennell2016life; @dennell2014hallam].
 
-The importance of the Movius' findings depend on the chronology of the artefacts he discovered.
-inferred from his geomorphological interpretations.
+The importance of the Movius' findings depend primarily on the chronology of the landscape he described, and the technology of the artefacts that he collected.
 Despite the prominence of the Movius Line in discussions of global patterns of human evolution, there has not been any new data from the location of Line since Movius' initial field work.
-In this paper we assess the current status of the
+Movius' interpretation of the Irrawaddy terraces and the technology of the artefacts he described remains untested by new data from his study area.
+In this paper we assess the current status of the Irrawaddy terraces and stratigraphy, and the technology of the artefacts found near the Irrawaddy in central Myanmar.
+We report on our recent surveys and field observations of the key archaeological locations that Movius visited in January of 1938.
 
@@
 
-Descriptions of the physical and cultural geography of the Irrawaddy terraces of central Myanmar have been appearing scholarly literature since Myanmar was attached to the British Empire in 1886.
-In the following section we review this literature to characterise this landscape.
+Descriptions of the physical and cultural geography of the Irrawaddy terraces of central Myanmar began appearing with increasing frequency in scholarly literature after Myanmar was attached to the British Empire in 1886.
+The Japanese occupation in 1942 disrupted geological research, and few publications appeared after the Second World War [@bender1983geology].
+In the following sections we review this literature to characterise this landscape.
 Then we review previous work on the archaeology of this landscape.
-Finally, we report on our recent surveys in the Chauk region, and discuss the implications for the archaeological significance of the region.
 
@@
 Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].
+
 In contrast to Dennell’s views, Norton et al.
@@
 The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts [@norton2006middle; @norton2008movius].
+
 Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life].
 Reasons for the lack of Acheulean technology are investigated in modern quaternary science [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius], including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
+
 The use of stone tools may have been influenced by certain situations, tasks or individuals.

### 2016-10-06 analysis/paper/paper.Rmd
more on movius

@@
 
-- Movius and contemporaries
+Since the 19th century, the focus of archaeological interest lay on the central part of the Irrawaddy River in central Myanmar.
+The first artefacts were discovered by Morris in 1930 in Upper Myanmar and later assigned to the “new” Anyathian culture from the Lower Palaeolithic in Southeast Asia.
+The Anyathian artefacts are mainly not found in situ but in the secondary deposits of the Pleistocene terrace gravels of the Irrawaddy River.
+The occurrence of only few handaxes is one main characteristic of the Anyanthian culture.
+The occupation of the Irrawaddy River terraces during the Pleistocene seems to be limited to the east bank of the River and the central part of Myanmar because the finds concentrate in the region around Chauk, Nyaung-U and Yenanyaung [@movius1944early; @deTerra1943].
 
-# Dennell is a major commentator recently
-Dennell is one of the more recent critics of the concept of the Movius Line.
-In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
-The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution.
-According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.
-Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].
+The three phases of the Early Anyathian are all dominated by a high degree of uniformity of the stone tools, probably due to raw material constraints (fossil wood and silicified tuff).
+Fossil wood is difficult to flake because it easily flakes parallel to the natural fibres of the wood.
+Most of the artefacts made from fossil wood are worked on only one plane; multiple flake scars are rare.
+Silicified tuff is more suitable for stone tool production due to its fine-grained homogenous texture and conchodial fractures, although it can be very porous.
+In total, 483 implements were found (261 from fossil wood, 220 from silicified tuff and 2 from quartzite).
+Implements made from fossil wood display a limited range of tools, including mostly handaxes, as well as choppers and chopping tools.
+Generally, stone tools made of silicified tuff are more massive than tools made from fossil wood.
+Choppers are dominant, whereas there are few handaxes [@deTerra1943].
 
-In contrast to Dennell’s views, Norton et al.
+In addition, 23 early Anyathian artefacts were found in situ in the lateritic gravel and gravel deposits of terrace 1 near Chauk on the Hill of Chinaungma.
+Movius (1943) assigned these implements to the first phase of the Early Anyathian and the beginning of the Middle Pleistocene as the oldest human artefacts found in the Irrawaddy Valley.
+Implements of phase 2 of the Early Anyathian show either a ferruginous crust, or are heavily rolled.
+The former implies an occupation of this area during the forming of the crust in a long interpluvial period.
+The latter suggests that these implements were derived from older deposits.
+Most of the Artefacts from the third phase of the Early Anyathian are very heavily rolled and are associated with the early Upper Pleistocene when the gravels were deposited [@deTerra1943].
+
+Norton et al.
 [-@norton2006middle] reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line.
@@
 
-Raw Material
-
 Movius [-@movius1944early] stated that the raw material used east of the Movius Line was often of low quality quartz and quartzite that would prevent the production of the same kind of bifaces [@dennell2016life; @lycett2010movius].
 However, handaxes from Zhoukoudian Locality 1, Chongokni and Kumpari (China) are bifacially worked and are made of quartz and quartzite river cobbles from this area [@lycett2010movius].
-
-Environmental Changes
 
@@
 However, even though bamboo cutting marks on bones can be distinguished from other cutting marks, there have been no findings of such so far [@lycett2010movius].
-
-Dispersal routes
 
@@
 
-Demographic and social transmission
-
 Recently, the model of demographic and social transmission has been proposed as a possible explanation for the diverse development of technologies in E/SE Asia and W/SW Asia.
@@
 Given the colonization of the whole of East Asia and the distance to Africa decreasing the population size, the population density may have been too low to maintain or establish more elaborate tool making techniques [@lycett2010demographic; @lycett2007there].
-
-[Figure demographic and social transmission?]
-
-Definition of “Acheulean”
 
@@
 Also, bifaces are not suitable as a time marker for specific cultures as they have been produced in different parts of the world over several 100 ka [@dennell2016life].
+
+Dennell is one of the more recent critics of the concept of the Movius Line.
+In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
+The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution.
+According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.
+Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].
 

### 2016-10-10 analysis/paper/paper.Rmd
new paragraph and sub-headings

@@
 
-Since the 19th century, the focus of archaeological interest lay on the central part of the Irrawaddy River in central Myanmar.
+### First archaeological exploration and the Movius Line
+
+Since the 19th century, archaeological interest was focussed on the central part of the Irrawaddy River in central Myanmar.
 The first artefacts were discovered by Morris in 1930 in Upper Myanmar and later assigned to the “new” Anyathian culture from the Lower Palaeolithic in Southeast Asia.
-The Anyathian artefacts are mainly not found in situ but in the secondary deposits of the Pleistocene terrace gravels of the Irrawaddy River.
-The occurrence of only few handaxes is one main characteristic of the Anyanthian culture.
+The Anyathian artefacts are mainly in the secondary deposits of the Pleistocene terrace gravels of the Irrawaddy River.
+The occurrence of only few handaxes is a main characteristic of Anyanthian assemblages.
 The occupation of the Irrawaddy River terraces during the Pleistocene seems to be limited to the east bank of the River and the central part of Myanmar because the finds concentrate in the region around Chauk, Nyaung-U and Yenanyaung [@movius1944early; @deTerra1943].
 
-The three phases of the Early Anyathian are all dominated by a high degree of uniformity of the stone tools, probably due to raw material constraints (fossil wood and silicified tuff).
+Due to raw material constraints (fossil wood and silicified tuff), the three phases of the Early Anyathian are dominated by a high degree of uniformity of the stone tools.
 Fossil wood is difficult to flake because it easily flakes parallel to the natural fibres of the wood.
@@
 In total, 483 implements were found (261 from fossil wood, 220 from silicified tuff and 2 from quartzite).
-Implements made from fossil wood display a limited range of tools, including mostly handaxes, as well as choppers and chopping tools.
-Generally, stone tools made of silicified tuff are more massive than tools made from fossil wood.
+Implements made from fossil wood include mostly handaxes, as well as choppers and chopping tools.
+Generally, stone tools made of silicified tuff are more massive compared to fossil wood.
 Choppers are dominant, whereas there are few handaxes [@deTerra1943].
 
-In addition, 23 early Anyathian artefacts were found in situ in the lateritic gravel and gravel deposits of terrace 1 near Chauk on the Hill of Chinaungma.
+In addition, 23 Early Anyathian artefacts were found in situ in the lateritic gravel and gravel deposits of terrace 1 near Chauk on the Hill of Chinaungma.
 Movius (1943) assigned these implements to the first phase of the Early Anyathian and the beginning of the Middle Pleistocene as the oldest human artefacts found in the Irrawaddy Valley.
@@
 
+Compared to the more complex technologies found in Africa, Europe, the Levant and western/southern Asia and the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line separating the Lower Palaeolithic into two cultures - the crude or non- stone tool-making cultures in E and SE Asia (e.g.
+Imjin/Hantan River Basin, Korea and Luonan Basin and Longyadong Cave, China) and northern China (e.g.
+Zhoukoudian) and the more sophisticated stone tool-making cultures in the Old World (e.g.
+Olduvai George Beds, Tanzania and Olorgesailie, Kenya) [@brumm2012biface2; @norton2006middle].
+
+### The Movius Line in modern quaternary science
+
+In modern quaternary science, the Movius Line is highly debated.
 Norton et al.

### 2016-10-10 analysis/paper/paper.Rmd
attempt to shorten the text

--- 2016-10-04 analysis/paper/paper.Rmd
@@
 
-- Movius and contemporaries
+### First archaeological exploration and the Movius Line
 
-# Dennell is a major commentator recently
-Dennell is one of the more recent critics of the concept of the Movius Line.
-In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
-The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution.
-According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.
-Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].
+Since the 19th century, archaeological interest was focussed on the central part of the Irrawaddy River in central Myanmar.
+The first artefacts were discovered by Morris in 1930 in Upper Myanmar later assigned to the Anyathian culture from the Lower Palaeolithic in Southeast Asia [@movius1948lowerand].
+Anyathian artefacts are mainly found in the secondary deposits of the Pleistocene terrace gravels of the Irrawaddy River, generally containing only few handaxes [@movius1948lowerand].
+The occupation of the Irrawaddy River terraces during the Pleistocene seems to be limited to the east bank of the River and the central part of Myanmar because the finds concentrate in the region around Chauk, Nyaung-U and Yenanyaung [@movius1944early; @deTerra1943].
 
-In contrast to Dennell’s views, Norton et al.
+Probably due to raw material constraints (fossil wood and silicified tuff), the three phases of the Early Anyathian are dominated by a high degree of uniformity of the stone tools [@movius1948lowerand].
+Fossil wood is difficult to flake because it easily flakes parallel to the natural fibres of the wood.
+Most of the artefacts made from fossil wood are worked on only one plane; multiple flake scars are rare.
+Silicified tuff is more suitable for stone tool production due to its fine-grained homogenous texture and conchodial fractures, although it can be very porous.
+In total, 483 implements were found (261 from fossil wood, 220 from silicified tuff and 2 from quartzite).
+Implements made from fossil wood include mostly handaxes, whereas stone tools made of silicified tuff are dominantly choppers [@deTerra1943; @movius1948lowerand].
+
+In addition, 23 Early Anyathian artefacts were found in situ in the lateritic gravel and gravel deposits of terrace 1 near Chauk on the Hill of Chinaungma.
+Movius [-@deTerra1943] assigned these implements to the first phase of the Early Anyathian and the beginning of the Middle Pleistocene as the oldest human artefacts found in the Irrawaddy Valley.
+Implements of phase 2 of the Early Anyathian show either a ferruginous crust, developping during an interpluvial period, or are heavily rolled, suggesting these implements have been derived from older deposits.
+Most of the Artefacts from the third phase of the Early Anyathian are very heavily rolled and are associated with the early Upper Pleistocene when the gravels were deposited [@deTerra1943].
+
+Comparing the more complex technologies found in Africa, Europe, the Levant and SW Asia and the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line separating the Lower Palaeolithic into two cultures - the crude or non- stone tool-making cultures in E and SE Asia (e.g.
+Imjin/Hantan River Basin, Korea and Luonan Basin and Longyadong Cave, China) and northern China (e.g.
+Zhoukoudian) and the more complex technologies in Africa, Europe and SW Asia (e.g.
+Olduvai George Beds, Tanzania and Olorgesailie, Kenya) [@brumm2012biface; @norton2006middle; @gao2014late; @dennell2014hallam].
+
+### The Movius Line in modern quaternary science
+
+In modern quaternary science, the Movius Line and its implications on human evolution in Southeast Asia are highly debated [e.g.
+@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius].
+Hutterer [-@hutterer1977reinterpreting] states only questionable geological associations between artefacts and the River terraces have been made and assumptions regarding the chronology of artefacts were mostly based on the degree of rolling [@dennell2014hallam].
+Norton et al.
 [-@norton2006middle] reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line.
@@
 
+Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma [@brumm2012biface; @lycett2010movius; @norton2006middle].
+These stone tools show a degree of standardization [@norton2006middle], included handaxes, cleavers, picks, knives, lanceolates and unifaces [@brumm2012biface] and are associated with hominins during the Pleistocene [@petraglia2008large].
+Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well [@brumm2012biface].
+The definition of the Acheulean is not always straightforward and bifaces have been produced in different parts of the world over several 100 ka [@dennell2016life].
+
 Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life].
-Reasons for the lack of Acheulean technology are investigated in modern quaternary science [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius], including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
+Reasons for the lack of Acheulean technology east of the Movius Line are recently more investigated, including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius].
 
 The use of stone tools may have been influenced by certain situations, tasks or individuals.
-Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times.
-Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool [@dennell2016life].
-
-Raw Material
+Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times [@dennell2016life].
+There may have been no need for intensive stone tool production, or a different material was used at some point.
+This theory is known as the “bamboo hypothesis” [@field2005assessment; @lycett2010movius; @schick1994movius] and states that early modern humans used bamboo instead of stones to make tools [@schick1994movius] leaving behind no archaeological record [@lycett2010movius].
 
@@
 
-Environmental Changes
-
 Whilst colonizing East Asia, early modern humans encountered barriers such as mountain ranges, river deltas, oceans and deserts that provide a range of resources (e.g.
@@
 However, the environment during the Middle Pleistocene was challenging in terms of changing climatic conditions and accompanying biogeographic transitions [@bar2001africa; @field2005assessment; @lycett2010movius; @schick1994movius].
-
-There may have been no need for intensive stone tool production, or a different material was used at some point.
-This theory is known as the “bamboo hypothesis” [@field2005assessment; @lycett2010movius; @schick1994movius] and states that early modern humans used bamboo instead of stones to make tools [@schick1994movius] leaving behind no archaeological record [@lycett2010movius].
-However, even though bamboo cutting marks on bones can be distinguished from other cutting marks, there have been no findings of such so far [@lycett2010movius].
-
-Dispersal routes
-
-The dispersal route of early modern humans from Africa to Asia and Australia is still debated among quaternary scientists.
 Based on GIS analyses, Field & Lahr [-@field2005assessment] and Field et al.
 [-@field2007southern] identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia.
-This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single].
-This route matches with some Middle Pleistocene archaeological sites in Africa, India (e.g.
-Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions.
-Of particular importance are the deposits at the Hiran Valley since they have been dated to 56-69ka, matching the proposed time for the Southern Dispersal.
+This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single] matching with some Middle Pleistocene archaeological sites in Africa, India (e.g.
+Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions dated to 56-69ka.
 A dispersal along the coasts and rivers seems reasonable because they provide sufficient resources.
-However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes QUELLE.
-
-Demographic and social transmission
+However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes.
 
@@
 
-[Figure demographic and social transmission?]
-
-Definition of “Acheulean”
-
-Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma [@brumm2012biface; @lycett2010movius; @norton2006middle].
-In contrary to Mode I technology, these stone tools show a degree of standardization [@norton2006middle] and were the main tools used by hominins during the Pleistocene [@petraglia2008large].
-The tools include handaxes, cleavers, picks, knives, lanceolates and unifaces [@brumm2012biface].
-Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well [@brumm2012biface].
-The definition of the Acheulean is not always straightforward, i.e.
-the assignment of finds can be influenced by the finding location east or west of the Movius Line QUELLE.
-Also, bifaces are not suitable as a time marker for specific cultures as they have been produced in different parts of the world over several 100 ka [@dennell2016life].
+Dennell is one of the more recent critics of the concept of the Movius Line.
+In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
+The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution [@dennell2014hallam; @dennell2016life].
+According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.
+Furthermore, because of the finding context, connections to other Middle Pleistocene Acheulean assemblages in Southwest Asia are difficult to demonstrate [@dennell2016life; @dennell2014hallam].
 

### 2016-10-11 analysis/paper/paper.Rmd
updated references, shortened text

@@
 Since the 19th century, archaeological interest was focussed on the central part of the Irrawaddy River in central Myanmar.
-The first artefacts were discovered by Morris in 1930 in Upper Myanmar later assigned to the Anyathian culture from the Lower Palaeolithic in Southeast Asia [@movius1948lowerand].
-Anyathian artefacts are mainly found in the secondary deposits of the Pleistocene terrace gravels of the Irrawaddy River, generally containing only few handaxes [@movius1948lowerand].
+The first artefacts were discovered by Morris in 1930 in Upper Myanmar later assigned to the Anyathian culture from the Lower Palaeolithic in Southeast Asia [@movius1948lower].
+Anyathian artefacts are mainly found in the secondary deposits of the Pleistocene terrace gravels of the Irrawaddy River, generally containing only few handaxes [@movius1948lower].
 The occupation of the Irrawaddy River terraces during the Pleistocene seems to be limited to the east bank of the River and the central part of Myanmar because the finds concentrate in the region around Chauk, Nyaung-U and Yenanyaung [@movius1944early; @deTerra1943].
 
-Probably due to raw material constraints (fossil wood and silicified tuff), the three phases of the Early Anyathian are dominated by a high degree of uniformity of the stone tools [@movius1948lowerand].
+Probably due to raw material constraints (fossil wood and silicified tuff), the three phases of the Early Anyathian are dominated by a high degree of uniformity of the stone tools [@movius1948lower].
 Fossil wood is difficult to flake because it easily flakes parallel to the natural fibres of the wood.
@@
 In total, 483 implements were found (261 from fossil wood, 220 from silicified tuff and 2 from quartzite).
-Implements made from fossil wood include mostly handaxes, whereas stone tools made of silicified tuff are dominantly choppers [@deTerra1943; @movius1948lowerand].
+Implements made from fossil wood include mostly handaxes, whereas stone tools made of silicified tuff are dominantly choppers [@deTerra1943; @movius1948lower].
 
@@
 
-Comparing the more complex technologies found in Africa, Europe, the Levant and SW Asia and the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line separating the Lower Palaeolithic into two cultures - the crude or non- stone tool-making cultures in E and SE Asia (e.g.
+Comparing the more complex technologies found in Africa, Europe, the Levant and SW Asia to the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line that separates the Lower Palaeolithic into two cultures - the crude or non- stone tool-making cultures in E and SE Asia (e.g.
 Imjin/Hantan River Basin, Korea and Luonan Basin and Longyadong Cave, China) and northern China (e.g.
@@
 
-Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life].
+Instead of assigning stone tools to certain technologies linked to different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life].
 Reasons for the lack of Acheulean technology east of the Movius Line are recently more investigated, including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius].
@@
 The use of stone tools may have been influenced by certain situations, tasks or individuals.
-Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times [@dennell2016life].
+Consequently, different individuals or groups could have used bifaces rarely, never or only at specific events or times [@dennell2016life].
 There may have been no need for intensive stone tool production, or a different material was used at some point.
@@
 [-@field2007southern] identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia.
-This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single] matching with some Middle Pleistocene archaeological sites in Africa, India (e.g.
+This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single; @marwick2009biogeography] matching with some Middle Pleistocene archaeological sites in Africa, India (e.g.
 Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions dated to 56-69ka.
 A dispersal along the coasts and rivers seems reasonable because they provide sufficient resources.
-However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes.
+However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes [@marwick2009biogeography].
 
@@
 The concept is based on the assumption that a certain effective population size is essential for developing traditions and technologies to be passed on to further generations [@lycett2010movius; @lycett2010demographic; @lycett2007there].
-Given the colonization of the whole of East Asia and the distance to Africa decreasing the population size, the population density may have been too low to maintain or establish more elaborate tool making techniques [@lycett2010demographic; @lycett2007there].
+Given the colonization of the whole of East Asia and the distance to Africa decreasing the population size, the population density may have been too low to maintain or establish more complex tool making techniques [@lycett2010demographic; @lycett2007there].
 
-Dennell is one of the more recent critics of the concept of the Movius Line.
-In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
-The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution [@dennell2014hallam; @dennell2016life].
+Dennell is one of the more recent critics of the concept of the Movius Line describing Movius' views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
+The drawing of a line to mark differences in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution [@dennell2014hallam; @dennell2016life].
 According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.

### 2016-10-11 analysis/paper/paper.Rmd
added some references

@@
 The concept is based on the assumption that a certain effective population size is essential for developing traditions and technologies to be passed on to further generations [@lycett2010movius; @lycett2010demographic; @lycett2007there].
-Given the colonization of the whole of East Asia and the distance to Africa decreasing the population size, the population density may have been too low to maintain or establish more complex tool making techniques [@lycett2010demographic; @lycett2007there].
+Given the colonization of the whole of East Asia and the distance to Africa decreasing the population size, the population density may have been too low to maintain or establish more complex tool making techniques [@lycett2010demographic; @lycett2007there; @dennell2014hallam].
 

### 2016-10-11 analysis/paper/paper.Rmd
more background text

--- 2016-10-10 analysis/paper/paper.Rmd
@@
 
-### First archaeological exploration and the Movius Line
+- Movius and contemporaries
 
-Since the 19th century, archaeological interest was focussed on the central part of the Irrawaddy River in central Myanmar.
-The first artefacts were discovered by Morris in 1930 in Upper Myanmar later assigned to the Anyathian culture from the Lower Palaeolithic in Southeast Asia [@movius1948lowerand].
-Anyathian artefacts are mainly found in the secondary deposits of the Pleistocene terrace gravels of the Irrawaddy River, generally containing only few handaxes [@movius1948lowerand].
-The occupation of the Irrawaddy River terraces during the Pleistocene seems to be limited to the east bank of the River and the central part of Myanmar because the finds concentrate in the region around Chauk, Nyaung-U and Yenanyaung [@movius1944early; @deTerra1943].
+# Dennell is a major commentator recently
+Dennell is one of the more recent critics of the concept of the Movius Line.
+In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
+The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution.
+According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.
+Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].
 
-Probably due to raw material constraints (fossil wood and silicified tuff), the three phases of the Early Anyathian are dominated by a high degree of uniformity of the stone tools [@movius1948lowerand].
-Fossil wood is difficult to flake because it easily flakes parallel to the natural fibres of the wood.
-Most of the artefacts made from fossil wood are worked on only one plane; multiple flake scars are rare.
-Silicified tuff is more suitable for stone tool production due to its fine-grained homogenous texture and conchodial fractures, although it can be very porous.
-In total, 483 implements were found (261 from fossil wood, 220 from silicified tuff and 2 from quartzite).
-Implements made from fossil wood include mostly handaxes, whereas stone tools made of silicified tuff are dominantly choppers [@deTerra1943; @movius1948lowerand].
-
-In addition, 23 Early Anyathian artefacts were found in situ in the lateritic gravel and gravel deposits of terrace 1 near Chauk on the Hill of Chinaungma.
-Movius [-@deTerra1943] assigned these implements to the first phase of the Early Anyathian and the beginning of the Middle Pleistocene as the oldest human artefacts found in the Irrawaddy Valley.
-Implements of phase 2 of the Early Anyathian show either a ferruginous crust, developping during an interpluvial period, or are heavily rolled, suggesting these implements have been derived from older deposits.
-Most of the Artefacts from the third phase of the Early Anyathian are very heavily rolled and are associated with the early Upper Pleistocene when the gravels were deposited [@deTerra1943].
-
-Comparing the more complex technologies found in Africa, Europe, the Levant and SW Asia and the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line separating the Lower Palaeolithic into two cultures - the crude or non- stone tool-making cultures in E and SE Asia (e.g.
-Imjin/Hantan River Basin, Korea and Luonan Basin and Longyadong Cave, China) and northern China (e.g.
-Zhoukoudian) and the more complex technologies in Africa, Europe and SW Asia (e.g.
-Olduvai George Beds, Tanzania and Olorgesailie, Kenya) [@brumm2012biface; @norton2006middle; @gao2014late; @dennell2014hallam].
-
-### The Movius Line in modern quaternary science
-
-In modern quaternary science, the Movius Line and its implications on human evolution in Southeast Asia are highly debated [e.g.
-@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius].
-Hutterer [-@hutterer1977reinterpreting] states only questionable geological associations between artefacts and the River terraces have been made and assumptions regarding the chronology of artefacts were mostly based on the degree of rolling [@dennell2014hallam].
-Norton et al.
+In contrast to Dennell’s views, Norton et al.
 [-@norton2006middle] reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line.
@@
 
-Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma [@brumm2012biface; @lycett2010movius; @norton2006middle].
-These stone tools show a degree of standardization [@norton2006middle], included handaxes, cleavers, picks, knives, lanceolates and unifaces [@brumm2012biface] and are associated with hominins during the Pleistocene [@petraglia2008large].
-Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well [@brumm2012biface].
-The definition of the Acheulean is not always straightforward and bifaces have been produced in different parts of the world over several 100 ka [@dennell2016life].
-
 Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life].
-Reasons for the lack of Acheulean technology east of the Movius Line are recently more investigated, including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius].
+Reasons for the lack of Acheulean technology are investigated in modern quaternary science [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius], including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
 
 The use of stone tools may have been influenced by certain situations, tasks or individuals.
-Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times [@dennell2016life].
-There may have been no need for intensive stone tool production, or a different material was used at some point.
-This theory is known as the “bamboo hypothesis” [@field2005assessment; @lycett2010movius; @schick1994movius] and states that early modern humans used bamboo instead of stones to make tools [@schick1994movius] leaving behind no archaeological record [@lycett2010movius].
+Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times.
+Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool [@dennell2016life].
+
+Raw Material
 
@@
 
+Environmental Changes
+
 Whilst colonizing East Asia, early modern humans encountered barriers such as mountain ranges, river deltas, oceans and deserts that provide a range of resources (e.g.
@@
 However, the environment during the Middle Pleistocene was challenging in terms of changing climatic conditions and accompanying biogeographic transitions [@bar2001africa; @field2005assessment; @lycett2010movius; @schick1994movius].
+
+There may have been no need for intensive stone tool production, or a different material was used at some point.
+This theory is known as the “bamboo hypothesis” [@field2005assessment; @lycett2010movius; @schick1994movius] and states that early modern humans used bamboo instead of stones to make tools [@schick1994movius] leaving behind no archaeological record [@lycett2010movius].
+However, even though bamboo cutting marks on bones can be distinguished from other cutting marks, there have been no findings of such so far [@lycett2010movius].
+
+Dispersal routes
+
+The dispersal route of early modern humans from Africa to Asia and Australia is still debated among quaternary scientists.
 Based on GIS analyses, Field & Lahr [-@field2005assessment] and Field et al.
 [-@field2007southern] identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia.
-This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single] matching with some Middle Pleistocene archaeological sites in Africa, India (e.g.
-Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions dated to 56-69ka.
+This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single].
+This route matches with some Middle Pleistocene archaeological sites in Africa, India (e.g.
+Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions.
+Of particular importance are the deposits at the Hiran Valley since they have been dated to 56-69ka, matching the proposed time for the Southern Dispersal.
 A dispersal along the coasts and rivers seems reasonable because they provide sufficient resources.
-However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes.
+However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes QUELLE.
+
+Demographic and social transmission
 
@@
 
-Dennell is one of the more recent critics of the concept of the Movius Line.
-In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life].
-The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution [@dennell2014hallam; @dennell2016life].
-According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene.
-Furthermore, because of the finding context, connections to other Middle Pleistocene Acheulean assemblages in Southwest Asia are difficult to demonstrate [@dennell2016life; @dennell2014hallam].
+[Figure demographic and social transmission?]
+
+Definition of “Acheulean”
+
+Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma [@brumm2012biface; @lycett2010movius; @norton2006middle].
+In contrary to Mode I technology, these stone tools show a degree of standardization [@norton2006middle] and were the main tools used by hominins during the Pleistocene [@petraglia2008large].
+The tools include handaxes, cleavers, picks, knives, lanceolates and unifaces [@brumm2012biface].
+Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well [@brumm2012biface].
+The definition of the Acheulean is not always straightforward, i.e.
+the assignment of finds can be influenced by the finding location east or west of the Movius Line QUELLE.
+Also, bifaces are not suitable as a time marker for specific cultures as they have been produced in different parts of the world over several 100 ka [@dennell2016life].
 

### 2016-10-12 analysis/paper/paper.Rmd
more on background geo

--- 2016-10-11 analysis/paper/paper.Rmd
@@
 ## Geomorphology of the Irrawady terraces
+
+There are three characteristic details of the terrace landforms: the red gravels that drape the surface, the fossil wood pieces found in these gravels, and the terraces
+
+One of the most distinctive features of the Irrawaddy terraces is the 'coarse red gravels', as they were described by Chhibber [-@chhibber1934geology].
+Chhibber describes the Irrawaddy River as meandering through the Lower Irrawaddy region, its course occasionaly constrained by sandstone cliffs.
+These cliffs are especially visible near Yenangyaung, and often capped by sheets of coarse red gravels.
+Chibber interprets the presence of these gravels as indicatative of previous hights of the river.
+Stamp [-@stamp1940irrawaddy, p.
+339] observed the 'thin, coarse red (partly lateritized) gravels' that characterise this landscape.
+
+The fossil wood fragments are typically within the size range of the background cobbles, but occasionally much larger trunk and branch sections are found.
+Chhibber notes that most of the fossil wood is identified Diptcrocarpoxylon burmense which grows abundantly in the region.
+Both monocotyledonous and dicotyledonous fossil wood are found in the deposits.
+A small percentage of the fossil wood is monocotyledoneae, with distinctive cordiform vascular bundles visible in the rock.
+The fossil wood is mostly siliceous, calcareous, ferruginous and carbonaceous.
+Most pieces are composed of chalcedony, microcrystalline silica, and opal.
+Rarely found are calcite, siderite, and quartz with nodules of iron and calcium [-@chhibber1934geology].
+Chhibber claims that the fossil wood results from colloidal silica dissolved in the waters that laid down the fossil-bearing deposit.
+
+Stamp recognised 'peneplaned surfaces at various levels' in the region, but is skeptical of de Terra's interpretations.
+Stamp writes that 'de Terra makes so many sweeping generalizations, some of which are obviously incorrect, and glosses over the most disputed difficulties of stratigraphical correlation, that one is forced to suggest that much more work is needed before his various conclusions can be accepted.'
 

### 2016-10-17 analysis/paper/paper.Rmd
update: Anyathian

@@
 
-Comparing the more complex technologies found in Africa, Europe, the Levant and SW Asia to the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line that separates the Lower Palaeolithic into two cultures - the crude or non- stone tool-making cultures in E and SE Asia (e.g.
+Comparing the more complex technologies found in Africa, Europe, the Levant and SW Asia to the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line that separates the Lower Palaeolithic into two cultures - the simple or non- stone tool-making cultures in E and SE Asia (e.g.
 Imjin/Hantan River Basin, Korea and Luonan Basin and Longyadong Cave, China) and northern China (e.g.
@@
 Olduvai George Beds, Tanzania and Olorgesailie, Kenya) [@brumm2012biface; @norton2006middle; @gao2014late; @dennell2014hallam].
+In addition, the term "Anyathian" is no longer used, instead the time span in question is referred to as the Middle Pleistocene.
 

### 2016-10-18 analysis/paper/paper.Rmd
Landscape observations

@@
 
-- our landscape observations
+### Landscape observations
+
+The landscape in the Lower Irrawaddy region is characterized by river terraces that are heavily incised.
+The uppermost terrace T1 is only sparsely preserved with one relict hill where the Chinthaungma monastery sits 70 m above the river level.
+Thick red gravel layers with a reddish-brown silty sand matrix overly the bedrock.
+
+Terrace 2 (T2) can be distinguished by the scarcity of vegetation compared to the lower terrace 3 (T3).
+Probably due to the very thin or absent soil layer on top pf the bedrock, only grass, shrubs and some scattered trees grow on T2.
+The terrain is severely incised and has a lot of gently rolling gullies and valleys.
+
+On T3, the vegetation is more prominent and there are fewer incisions.
+The red gravel layer is not as dominant as in T1, whereas the soil layer seems to be more pronounced.
+There are some filled fields and a golf course that was created in 1947 for the local oil workers.
+Because of its lower location, the gravels may be colluvial redposits from T1, although it could be an alluvial terrace.
+
 - descriptions of existing cuttings that we visited

### 2016-10-18 analysis/paper/paper.Rmd
more landscape observation

@@
 
+According to Movius, two more terraces can be spotted from the river.
+After an intensive inspection, terrace 4 (T4) may be a cliff close to the river (base at 46 m) with a slight elevation  compared to the river (41 m).
+In contrast to the other terraces, the red gravel layer is absent on T4.
+However, the bedrock is overlain by  coarse gravels in a reddish brown sandy silt matrix (5 cm) that is followed by a reddish brown fine grained sandy silt with occasional small gravel clasts, as well as small carbonate modules.
+
 - descriptions of existing cuttings that we visited

### 2016-10-18 analysis/paper/paper.Rmd
description cuttings

@@
 
+In general, the cutting consist of 4 layers and a thin soil profile of about 5 cm.
+The uppermost layer is made from brownish red poorly sorted gravel that contains medium to well-rounded cobbles and pebbles.
+The slightly silty sandy matrix contains an increasing amount of gravel towards the top.
+The cobbles are weathered and the larger ones are made from fossil wood.
+There is no bedding, sorting or imbrication visible.
+Fine roots penetrate the profile throughout.
+The lower contact with the following layer is moderately sharp.
+The gravel layer is followed by a brownish red silty sand that is well rounded with no obvious bedding.
+There occasional lighter patches that could indicate decayed pebbles.
+The contact with the next layer is fairly sharp.
+Between the bedrock and the red sand layer there is angular weathered bedrock intertwined with a dark band, possibly either a weathered palaeosurface or a weathered intrusion of host rock.
+The bedrock itself is medium to fine-grained light grey and brown red sand including some silt.
+The red band could be filling from above.
+A few isolated roots are visible.
+
 - descriptions of existing cuttings that we visited

### 2016-10-18 analysis/paper/paper.Rmd
update cuttings
sketches

--- 2016-10-18 analysis/paper/paper.Rmd
@@
 
-Comparing the more complex technologies found in Africa, Europe, the Levant and SW Asia to the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line that separates the Lower Palaeolithic into two cultures - the simple or non- stone tool-making cultures in E and SE Asia (e.g.
+Comparing the more complex technologies found in Africa, Europe, the Levant and SW Asia to the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line that separates the Lower Palaeolithic into two cultures - the crude or non- stone tool-making cultures in E and SE Asia (e.g.
 Imjin/Hantan River Basin, Korea and Luonan Basin and Longyadong Cave, China) and northern China (e.g.
@@
 Olduvai George Beds, Tanzania and Olorgesailie, Kenya) [@brumm2012biface; @norton2006middle; @gao2014late; @dennell2014hallam].
-In addition, the term "Anyathian" is no longer used, instead the time span in question is referred to as the Middle Pleistocene.
 
@@
 
-### Landscape observations
-
-The landscape in the Lower Irrawaddy region is characterized by river terraces that are heavily incised.
-The uppermost terrace T1 is only sparsely preserved with one relict hill where the Chinthaungma monastery sits 70 m above the river level.
-Thick red gravel layers with a reddish-brown silty sand matrix overly the bedrock.
-
-Terrace 2 (T2) can be distinguished by the scarcity of vegetation compared to the lower terrace 3 (T3).
-Probably due to the very thin or absent soil layer on top pf the bedrock, only grass, shrubs and some scattered trees grow on T2.
-The terrain is severely incised and has a lot of gently rolling gullies and valleys.
-
-On T3, the vegetation is more prominent and there are fewer incisions.
-The red gravel layer is not as dominant as in T1, whereas the soil layer seems to be more pronounced.
-There are some filled fields and a golf course that was created in 1947 for the local oil workers.
-Because of its lower location, the gravels may be colluvial redposits from T1, although it could be an alluvial terrace.
-
+- our landscape observations
 - descriptions of existing cuttings that we visited
@@
 ## References
+
+knitr::current_input()

### 2016-10-18 analysis/paper/paper.Rmd
more on geo background

--- 2016-10-18 analysis/paper/paper.Rmd
@@
 These cliffs are especially visible near Yenangyaung, and often capped by sheets of coarse red gravels.
-Chibber interprets the presence of these gravels as indicatative of previous hights of the river.
+Chibber interprets the presence of these gravels as indicatative of previous heights of the river.
 Stamp [-@stamp1940irrawaddy, p.
-339] observed the 'thin, coarse red (partly lateritized) gravels' that characterise this landscape.
+339] also observed the 'thin, coarse red (partly lateritized) gravels' that characterise this landscape.
+He traces their origin to the Arnkun Yoma, 150 km to the north west, based on the composition and morphology of the cobbles.
+Chhibber and Stamp both made a broad correlation between the red gravels of the Irrawaddy and the Siwaliks of India based on similarities in the large vertebrate fossil taxa, for example, _Mastodon_ sp., _Stegodon_ sp., and _Hippoputamus_ sp.
+This implies a Pliocene age for Irrawaddy deposits.
 
-The fossil wood fragments are typically within the size range of the background cobbles, but occasionally much larger trunk and branch sections are found.
-Chhibber notes that most of the fossil wood is identified Diptcrocarpoxylon burmense which grows abundantly in the region.
+de Terra et al.
+[-@deTerra1943] calls these deposits 'Lateritic Gravel' and claim that the Lateritic Gravel of the Chauk area are related to the Uru Boulder Conglomerate of Northern Burma (400 km north of Chauk), and the uppermost Siwalik Boulders of Northwestern India.
+The correlation is based on the presence of greenstone, gneissic rocks and sandstone in the gravels.
+This correlation with India is the basis for their determination that the Irrawaddy terraces formed during the Plio-Pleistocene.
+Dennell [-@dennell2014hallam] has reviewed this claim in detail, and found no convincing evidence for a connection between India and these Irrawaddy gravels, noting that greenstone, gneissic rocks and sandstone are all present in the metamorphic belt of central Burma that extends from the Myityina district in upper Myanmar, to south of Mandalay.
+
+The fossil wood fragments found in these gravels are typically within the size range of the background cobbles, but occasionally much larger trunk and branch sections are found.
+Chhibber notes that most of the fossil wood is identified _Diptcrocarpoxylon burmense_ which grew abundantly in the region at the time of his survey.
 Both monocotyledonous and dicotyledonous fossil wood are found in the deposits.
@@
 Rarely found are calcite, siderite, and quartz with nodules of iron and calcium [-@chhibber1934geology].
-Chhibber claims that the fossil wood results from colloidal silica dissolved in the waters that laid down the fossil-bearing deposit.
+Chhibber discussed a range of previous theories about remote origins of the fossil wood, but finds them unconvinving because of the distribution of fossil wood is limited to the Irrawaddy valley.
+He concludes that the most likely origin is weathering combined with fresh-water desert conditions to produce colloidal silica that dissolved in the waters that laid down the gravel deposit in the Irrawaddy valley region.
+Chhibber's primary evidence for this local origin is the identification of Lieseqamq rings in the fossil wood, a characteristic of chemical precipitation reactions.
+
+Perhaps the most controversial geomorhological detail of the Irrawaddy is the identification and chronology of the terraces. ...de Terra... ...Chhibber...
 
@@
 Stamp writes that 'de Terra makes so many sweeping generalizations, some of which are obviously incorrect, and glosses over the most disputed difficulties of stratigraphical correlation, that one is forced to suggest that much more work is needed before his various conclusions can be accepted.'
-
-Rep
-
-chhibber1934geology
-
-- survey early literature on red earth deposits
 
@@
 
-In general, the cutting consist of 4 layers and a thin soil profile of about 5 cm.
+In general, the cuttings consist of 4 layers and a thin soil profile of about 5 cm.
 The uppermost layer is made from brownish red poorly sorted gravel that contains medium to well-rounded cobbles and pebbles.

### 2016-10-20 analysis/paper/paper.Rmd
further organisation
Update: landform and west side of the river

@@
 
-## Survey data
-
-```{r}
-# analysis of survey and GPS data in here
-```
-
 ### Landscape observations
 
-The landscape in the Lower Irrawaddy region is characterized by river terraces that are heavily incised.
-The uppermost terrace T1 is only sparsely preserved with one relict hill where the Chinthaungma monastery sits 70 m above the river level.
+The landscape in the Lower Irrawaddy region is characterized by a series of landforms on different levels on the eastern side of the river.
+There are sand cliffs that are heavily incised.
+On the west bank of the river, an old maeander of the Irrawaddy, but no terrace-like forms can be seen apart from the remnants of a plateau or peneplaned surface further inland.
+Another landscape feature is the geological fault zone east of Chauk, marking the base of the Shan scarp.
+
+The uppermost landform (L1) is only sparsely preserved with one relict hill where the Chinthaungma monastery sits 70 m above the river level.
 Thick red gravel layers with a reddish-brown silty sand matrix overly the bedrock.
 
-Terrace 2 (T2) can be distinguished by the scarcity of vegetation compared to the lower terrace 3 (T3).
-Probably due to the very thin or absent soil layer on top pf the bedrock, only grass, shrubs and some scattered trees grow on T2.
+Landform 2 (L2) can be distinguished by the scarcity of vegetation compared to the lower landform 3 (L3).
+Probably due to the very thin or absent soil layer on top pf the bedrock, only grass, shrubs and some scattered trees grow on L2.
 The terrain is severely incised and has a lot of gently rolling gullies and valleys.
 
-On T3, the vegetation is more prominent and there are fewer incisions.
-The red gravel layer is not as dominant as in T1, whereas the soil layer seems to be more pronounced.
+On L3, the vegetation is more prominent and there are fewer incisions.
+The red gravel layer is not as dominant as on L1, whereas the soil layer seems to be more pronounced.
 There are some filled fields and a golf course that was created in 1947 for the local oil workers.
-Because of its lower location, the gravels may be colluvial redposits from T1, although it could be an alluvial terrace.
+Because of its lower location, the gravels may be colluvial redposits from L1, although it could be an alluvial terrace.
 
-According to Movius, two more terraces can be spotted from the river.
-After an intensive inspection, terrace 4 (T4) may be a cliff close to the river (base at 46 m) with a slight elevation  compared to the river (41 m).
-In contrast to the other terraces, the red gravel layer is absent on T4.
-However, the bedrock is overlain by coarse gravels in a reddish brown sandy silt matrix (5 cm) that is followed by a reddish brown fine grained sandy silt with occasional small gravel clasts, as well as small carbonate modules.
+According to Movius, two more levels can be spotted from the river.
+After an intensive inspection, the fourth landform (L4) may be a cliff close to the river (base at 46 m) with a slight elevation compared to the river (41 m).
+In contrast to the other levels, the red gravel layer is absent on L4.
+However, the bedrock is overlain by coarse gravels in a reddish brown sandy silt matrix (5 cm) that is followed by reddish brown fine grained sandy silt with occasional small gravel clasts, as well as small carbonate modules.
+
+However, these landforms cannot be described as terraces for certain.
+There are indications that it might be terraces as elucidated by Movius, e.g.
+the gravels found on L1 and L3.
+But explicit terraces could not be identified in the field, especially due to the absence of terraces on the west side of the Irrawaddy River.
+More intensive cultivation and a very disturbed landscape could be reasons for that.
 
@@
 
-- descriptions of existing cuttings that we visited
-- GPS data from transect
-- surface finds of artefacts by KK
+## Survey data
+
+```{r}
+# analysis of survey and GPS data in here
+```
+
+## Surface finds of stone artefacts
 

### 2017-01-11 analysis/paper/paper.Rmd
inserted picture of landscape
minor changes

@@
 Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma [@brumm2012biface; @lycett2010movius; @norton2006middle].
-These stone tools show a degree of standardization [@norton2006middle], included handaxes, cleavers, picks, knives, lanceolates and unifaces [@brumm2012biface] and are associated with hominins during the Pleistocene [@petraglia2008large].
+These stone tools show a degree of standardization [@norton2006middle], including handaxes, cleavers, picks, knives, lanceolates and unifaces [@brumm2012biface] and are associated with hominins during the Pleistocene [@petraglia2008large].
 Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well [@brumm2012biface].
@@
 Consequently, different individuals or groups could have used bifaces rarely, never or only at specific events or times [@dennell2016life].
-There may have been no need for intensive stone tool production, or a different material was used at some point.
+There may have been no need for intensive stone tool production, or a different material was used, e.g.
+bamboo.
 This theory is known as the “bamboo hypothesis” [@field2005assessment; @lycett2010movius; @schick1994movius] and states that early modern humans used bamboo instead of stones to make tools [@schick1994movius] leaving behind no archaeological record [@lycett2010movius].
@@
 Another landscape feature is the geological fault zone east of Chauk, marking the base of the Shan scarp.
+
+![Landscape around the middle part of the Irrawaddy River near Chauk.
+View from the Chinthaungma monastery towards the west side of the river.](C:/Users/Maria/Documents/PhD Wollongong/Myanmar/Chauk/chaukgeomorph/P6200093.JPG)
 

### 2017-01-16 analysis/paper/paper.Rmd
Figure 1

@@
 
-![Landscape around the middle part of the Irrawaddy River near Chauk.
-View from the Chinthaungma monastery towards the west side of the river.](C:/Users/Maria/Documents/PhD Wollongong/Myanmar/Chauk/chaukgeomorph/P6200093.JPG)
+```{r LandscapeChauk, fig.cap="Landscape around the middle part of the Irrawaddy River near Chauk. View from the Chinthaungma monastery towards the west side of the river."}
+knitr::include_graphics("../figures/Landscape_Chauk.JPG")
+```
 
@@
 Landform 2 (L2) can be distinguished by the scarcity of vegetation compared to the lower landform 3 (L3).
-Probably due to the very thin or absent soil layer on top pf the bedrock, only grass, shrubs and some scattered trees grow on L2.
+Probably due to the very thin or absent soil layer on top of the bedrock, only grass, shrubs and some scattered trees grow on L2 (Figure \@ref(fig:LandscapeChauk)).
 The terrain is severely incised and has a lot of gently rolling gullies and valleys.
